---
title: "SCM02_walkthrough"
format: html
editor_options: 
  chunk_output_type: console
---

## Workspace set-up

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(mgcv)
library(dagitty)
library(igraph)
library(broom)
library(furrr)
library(future)
library(gratia)
library(DHARMa)
```


## Generate the fake dataset for this example (R code)

```{r}
set.seed(20260408)

n_site <- 4
n_obs  <- 5
n_rep  <- 10
n      <- n_site * n_obs * n_rep

site <- factor(rep(paste0("site_", seq_len(n_site)), each = n_obs * n_rep))
obs  <- factor(rep(rep(paste0("obs_", seq_len(n_obs)), each = n_rep), times = n_site))

b_site_x  <- rnorm(n_site, 0, 0.55)
b_obs_x   <- rnorm(n_obs, 0, 0.45)

b_site_m1 <- rnorm(n_site, 0, 0.35)
b_obs_m1  <- rnorm(n_obs, 0, 0.30)

b_site_m2 <- rnorm(n_site, 0, 0.35)
b_obs_m2  <- rnorm(n_obs, 0, 0.30)

b_site_y  <- rnorm(n_site, 0, 0.70)
b_obs_y   <- rnorm(n_obs, 0, 0.55)

C <- rnorm(n, mean = 0, sd = 1.1)

X <- 0.85 * C +
  b_site_x[site] +
  b_obs_x[obs] +
  rnorm(n, 0, 1.0)

M1 <- 1.0 * sin(X / 1.3) +
  0.55 * C +
  b_site_m1[site] +
  b_obs_m1[obs] +
  rnorm(n, 0, 0.8)

M2 <- 0.60 * X +
  0.85 * M1 +
  0.18 * C +
  b_site_m2[site] +
  b_obs_m2[obs] +
  rnorm(n, 0, 0.9)

Y <- 1.15 * X +
  1.10 * M2 -
  0.42 * (M2^2) +
  0.70 * C +
  b_site_y[site] +
  b_obs_y[obs] +
  rt(n, df = 6) * 0.9

dat <- tibble(
  site = site,
  obs  = obs,
  C    = C,
  X    = X,
  M1   = M1,
  M2   = M2,
  Y    = Y
)

glimpse(dat)
```

---

## Submodel for `X`

## Error distributions for `X`

```{r}
x_error01 <- gam(
  X ~ C,
  data = dat,
  family = gaussian(),
  method = "REML"
)

x_error02 <- gam(
 X ~ C,
  data = dat,
  family = scat(),
  method = "REML"
)

tibble(
  model = c("X: Gaussian", "X: scat"),
  AIC = c(AIC(x_error01), AIC(x_error02))
) |>
  arrange(AIC) |>
  mutate(delta_AIC = AIC - min(AIC))
```

## Random effects structure for 'X'

```{r}
x_re01 <- gam(
  X ~ C + s(site, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

x_re02 <- gam(
  X ~ C + s(obs, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

x_re03 <- gam(
  X ~ C + s(site, bs = "re") + s(obs, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

tibble(
  model = c("X: site RE", "X: obs RE", "X: crossed REs"),
  AIC = c(AIC(x_re01), AIC(x_re02), AIC(x_re03))
) |>
  arrange(AIC) |>
  mutate(delta_AIC = AIC - min(AIC))
```

## Functional form for `X`

```{r}
x_form01 <- gam(
  X ~ C + s(site, bs = "re") + s(obs, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

x_form02 <- gam(
  X ~ s(C) + s(site, bs = "re") + s(obs, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

tibble(
  model = c("X: linear", "X: smooth X"),
  AIC = c(AIC(x_form01), AIC(x_form02))
) |>
  arrange(AIC) |>
  mutate(delta_AIC = AIC - min(AIC))

```

---

# Submodel for `M1`

## Error distribution for `M1`

```{r}
m1_error01 <- gam(
  M1 ~ X + C,
  data = dat,
  family = gaussian(),
  method = "REML"
)

m1_error02 <- gam(
  M1 ~ X + C,
  data = dat,
  family = scat(),
  method = "REML"
)

tibble(
  model = c("M1: gaussian", "M1: scaled-t"),
  AIC = c(AIC(m1_error01), AIC(m1_error02))
) |>
  arrange(AIC) |>
  mutate(delta_AIC = AIC - min(AIC))

```


## Random effects structure for `M1`

```{r}
m1_re01 <- gam(
  M1 ~ X + C + s(site, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

m1_re02 <- gam(
  M1 ~ X + C + s(obs, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

m1_re03 <- gam(
  M1 ~ X + C + s(site, bs = "re") + s(obs, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

tibble(
  model = c("M1: site RE", "M1: obs RE", "M1: crossed REs"),
  AIC = c(AIC(m1_re01), AIC(m1_re02), AIC(m1_re03))
) |>
  arrange(AIC) |>
  mutate(delta_AIC = AIC - min(AIC))
```
:::

### Function form for 'M1'

```{r}
m1_form01 <- gam(
  M1 ~ X + C + s(site, bs = "re") + s(obs, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

m1_form02 <- gam(
  M1 ~ s(X) + C + s(site, bs = "re") + s(obs, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

m1_form03 <- gam(
  M1 ~ X + s(C) + s(site, bs = "re") + s(obs, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

m1_form04 <- gam(
  M1 ~ s(X) + s(C) + s(site, bs = "re") + s(obs, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

tibble(
  model = c("M1: all linear", "M1: smooth X", "M1: smooth C", "M1: smooth both"),
  AIC = c(AIC(m1_form01), AIC(m1_form02), AIC(m1_form03), AIC(m1_form04))
) |>
  arrange(AIC) |>
  mutate(delta_AIC = AIC - min(AIC))
```

---

## Submodel for `M2`

### Error distribution for `M2`

```{r}
m2_error01 <- gam(
  M2 ~ X + M1,
  data = dat,
  family = gaussian(),
  method = "REML"
)

m2_error02 <- gam(
  M2 ~ X + M1,
  data = dat,
  family = scat(),
  method = "REML"
)

tibble(
  model = c("M2: gaussian", "M2: scat"),
  AIC = c(AIC(m2_error01), AIC(m2_error02))
) |>
  arrange(AIC) |>
  mutate(delta_AIC = AIC - min(AIC))
```

### Random effects structure for `M2`

```{r}
m2_re01 <- gam(
  M2 ~ X + M1 + s(site, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

m2_re02 <- gam(
  M2 ~ X + M1 + s(obs, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

m2_re03 <- gam(
  M2 ~ X + M1 + s(site, bs = "re") + s(obs, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

tibble(
  model = c("M2: site RE", "M2: obs RE", "M2: crossed REs"),
  AIC = c(AIC(m2_re01), AIC(m2_re02), AIC(m2_re03))
) |>
  arrange(AIC) |>
  mutate(delta_AIC = AIC - min(AIC))
```

### Functional form for `M2`

```{r}
m2_form01 <- gam(
  M2 ~ X + M1 + s(site, bs = "re") + s(obs, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

m2_form02 <- gam(
  M2 ~ s(X) + M1 + s(site, bs = "re") + s(obs, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

m2_form03 <- gam(
  M2 ~ X + s(M1) + s(site, bs = "re") + s(obs, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

m2_form04 <- gam(
  M2 ~ s(X) + s(M1) + s(site, bs = "re") + s(obs, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

tibble(
  model = c("M2: all linear", "M2: smooth X", "M2: smooth C", "M2: smooth both"),
  AIC = c(AIC(m2_form01), AIC(m2_form02), AIC(m2_form03), AIC(m2_form04))
) |>
  arrange(AIC) |>
  mutate(delta_AIC = AIC - min(AIC))
```


# The part uses ALL models in the second part of the SCM tutorial


```{r}
# function to return the most common level of a factor
mode_factor <- function(x) {
  factor(names(which.max(table(x))), levels = levels(x))
}

# construct the baseline prediction dataset
new_base <- tibble(
  C = dat$C,
  site = mode_factor(dat$site),
  obs  = mode_factor(dat$obs)
)
```



### Substep 7c: Specify the intervention(s)

```{r}
x0 <- quantile(dat$X, 0.25)
x1 <- quantile(dat$X, 0.75)

c(x0 = x0, x1 = x1)
```



### Substep 7d: Predict outcomes under each intervention to obtain the *total effect*


```{r}
predict_under_do_x <- function(x_value, base_dat) {

  dat_x <- base_dat |>
    mutate(X = x_value)
  m1_hat <- predict(mod_m1, newdata = dat_x, type = "response")

  dat_m2 <- dat_x |>
    mutate(M1 = m1_hat)
  m2_hat <- predict(mod_m2, newdata = dat_m2, type = "response")

  dat_y <- dat_m2 |>
    mutate(M2 = m2_hat)
  y_hat <- predict(mod_y, newdata = dat_y, type = "response")

  y_hat
}
```


### Substep 7e: Compute the *controlled direct effect* by fixing mediator(s)


```{r}
predict_cde <- function(x_treat, x_mediator_ref, base_dat) {

  dat_xref <- base_dat |>
    mutate(X = x_mediator_ref)
  m1_ref <- predict(mod_m1, newdata = dat_xref, type = "response")

  dat_m2ref <- dat_xref |>
    mutate(M1 = m1_ref)
  m2_ref <- predict(mod_m2, newdata = dat_m2ref, type = "response")

  dat_y <- base_dat |>
    mutate(
      X = x_treat,
      M2 = m2_ref
    )
  y_hat <- predict(mod_y, newdata = dat_y, type = "response")

  y_hat
}
```

## Substep 7f: Decompose the *total effect* into *direct* and *indirect* components


```{r}
y_do_x1 <- predict_under_do_x(x1, new_base)
y_do_x0 <- predict_under_do_x(x0, new_base)

te_hat <- mean(y_do_x1 - y_do_x0)

# controlled direct effect: hold mediator pathway at its x0 level
cde_x1_x0 <- mean(
  predict_cde(x_treat = x1, x_mediator_ref = x0, base_dat = new_base) -
    predict_cde(x_treat = x0, x_mediator_ref = x0, base_dat = new_base)
)

ie_hat <- te_hat - cde_x1_x0

tibble(
  effect = c("Total effect", "Controlled direct effect", "Indirect effect"),
  estimate = c(te_hat, cde_x1_x0, ie_hat)
)
```



```{r}

boot_decomp <- function(idx, data, x0, x1) {
  
  # idx = the indices for the resampled rows
  d <- data[idx, ]

  ref_site_b <- mode_factor(d$site)
  ref_obs_b  <- mode_factor(d$obs)

  new_base_b <- tibble(
    C = d$C,
    site = factor(ref_site_b, levels = levels(data$site)),
    obs  = factor(ref_obs_b,  levels = levels(data$obs))
  )

  # Now run each dataset through the specific order of the causal models.
  mod_m1_b <- gam(
    M1 ~ s(X) + s(C) + s(site, bs = "re") + s(obs, bs = "re"),
    data = d,
    family = gaussian(),
    method = "REML"
  )

  mod_m2_b <- gam(
    M2 ~ s(X) + s(M1) + s(site, bs = "re") + s(obs, bs = "re"),
    data = d,
    family = gaussian(),
    method = "REML"
  )

  mod_y_b <- gam(
    Y ~ s(X) + s(M2) + s(C) + s(site, bs = "re") + s(obs, bs = "re"),
    data = d,
    family = scat(),
    method = "REML"
  )

  # predict total effect (TE)
  predict_under_do_x_b <- function(x_value, base_dat) {
    dat_x <- base_dat |> mutate(X = x_value)
    m1_hat <- predict(mod_m1_b, newdata = dat_x, type = "response")
    dat_m2 <- dat_x |> mutate(M1 = m1_hat)
    m2_hat <- predict(mod_m2_b, newdata = dat_m2, type = "response")
    dat_y <- dat_m2 |> mutate(M2 = m2_hat)
    predict(mod_y_b, newdata = dat_y, type = "response")
  }

  # predict controlled direct effect (CDE)
  predict_cde_b <- function(x_treat, x_mediator_ref, base_dat) {
    dat_xref <- base_dat |> mutate(X = x_mediator_ref)
    m1_ref <- predict(mod_m1_b, newdata = dat_xref, type = "response")
    dat_m2ref <- dat_xref |> mutate(M1 = m1_ref)
    m2_ref <- predict(mod_m2_b, newdata = dat_m2ref, type = "response")
    dat_y <- base_dat |> mutate(X = x_treat, M2 = m2_ref)
    predict(mod_y_b, newdata = dat_y, type = "response")
  }

  # Differences are the counterfactual contrasts
  te <- mean(predict_under_do_x_b(x1, new_base_b) - predict_under_do_x_b(x0, new_base_b))

  de <- mean(
    predict_cde_b(x_treat = x1, x_mediator_ref = x0, base_dat = new_base_b) -
      predict_cde_b(x_treat = x0, x_mediator_ref = x0, base_dat = new_base_b)
  )

  # values are not in any order, so we can simply take the difference to get the distribution of IE
  ie <- te - de

  # create output data for this
  tibble(te = te, de = de, ie = ie)
}
```

Now, we can run the bootstrap.

```{r}

suppressWarnings(plan(multisession))

set.seed(99)
B <- 200
boot_ids <- replicate(B, sample(seq_len(nrow(dat)), replace = TRUE), simplify = FALSE)

boot_res <- future_map_dfr(boot_ids, boot_decomp, data = dat, x0 = x0, x1 = x1, .progress = TRUE)

boot_summary <- tibble(
  effect = c("Total effect", "Controlled direct effect", "Indirect effect"),
  estimate = c(mean(boot_res$te), mean(boot_res$de), mean(boot_res$ie)),
  lwr = c(quantile(boot_res$te, 0.025), quantile(boot_res$de, 0.025), quantile(boot_res$ie, 0.025)),
  upr = c(quantile(boot_res$te, 0.975), quantile(boot_res$de, 0.975), quantile(boot_res$ie, 0.975))
)

boot_summary
```

## Step 8: Visualize the decomposition

```{r}
boot_summary |>
  ggplot(aes(x = effect, y = estimate, ymin = lwr, ymax = upr)) +
  geom_pointrange() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Estimated causal contrast",
    title = "Total, direct, and indirect effects from the fitted SCM"
  )
```