---
title: "SCM: Total Effects"
format:
  html:
    toc: true
    toc-depth: 3
    mermaid:
      theme: forest
---

## Introduction

This tutorial shows how to estimate **total**, **direct**, and **indirect** effects in a nonlinear structural causal model (SCM) using **frequentist methods only** (i.e. non Bayesian), with `mgcv::gam()` as the primary modeling tool. The example is built to be scientifically defensible rather than merely predictive. Like you would do for a scientific manuscript, this tutorial:

1. states the causal DAG explicitly
2. uses `dagitty` to define the adjustment set for the total effect
3. simulates data from a known SCM with both linear and nonlinear edges
4. includes two crossed random effects (`site` and `observer`)
5. estimates direct and indirect effects by `g-computation`, not by multiplying coefficients

Again, note that the first workflow you will be exposed to is entirely frequentist (something with which you are already familiar). 

* `mgcv::gam()` for structural equations,
* AIC for pathway model (i.e. sub-model) comparison where appropriate
* bootstrap resampling for uncertainty in `g-computation`

This section introduces a few key ideas using simple math. Do not worry. We will walk through the process step by step. We will cover the total effect, the direct effect (controlling for both mediators and confounders), and the indirect effect. Although causal decomposition (e.g., via g-computation) allows us to separate direct and indirect effects, we begin with the total effect and then, in the next tutorial, will cover formal causal decomposition. This is typically the first model you fit when addressing a broad causal question, before examining specific pathways or mediation. 

::: callout-warning
## Centering and scaling of predictors is usually recommended.

In usual analyses, it is generally advisable for all *predictors that appear on the right-hand side of equations* to be centered and scaled before fitting the final structural equations. This is done for the following practical reasons:

* it makes optimization and estimation more stable
* it puts predictors on comparable numerical ranges
* it reduces problems when some predictors are measured in very different units (this is the main reason)
* it makes intervention values easier to define (oh wait, this is the main reason)
* and it often helps smooth terms and random effects behave more predictably

The final outcome `Y` is kept on its **raw scale** so the estimated causal contrasts remain interpretable in the original outcome units. 

That is the key rule here:

> Standardize predictors for estimation convenience; keep the final outcome on its natural scale for interpretation.

:::

## The scientific setup for this tutorial's question

For our example, let us consider the following example (not a jackalope example). Suppose our primary goal is to test for the causal effect of an exposure `X` on a continuous outcome `Y`. 

A baseline confounder `C` affects both `X` and `Y`, and also affects mediator `M1`.
The effect of `X` on `Y` is partly direct and partly transmitted through two mediators:

* `X -> M1 -> M2 -> Y`
* `X -> M2 -> Y`

We also allow repeated observations arising from two *crossed grouping factors* (which become crossed random effects in our model):

* `site`
* `observer`

These are not nested. Each observer can appear at multiple sites, and each site can be visited by multiple observers. Crossed --rather than nested-- random effects are used in this example just to make the AIC comparisons a bit more interesting. 

## Constructing a causal Directed Acyclic Graph (DAG) of our system

```{mermaid}

flowchart LR

    C[C]
    X[X]
    M1[M1]
    M2[M2]
    Y[Y]

    %% Edges
    C --> X
    C --> M1
    C --> Y
    X --> M1

    X --> Y
    M1 --> M2
    X --> M2
    M2 --> Y

    %% Styles (fully editable)
    style C fill:#FACECE,stroke:none,color:#000000,font-size:16px
    style X fill:#D6EAF8,stroke:none,color:#000000,font-size:16px
    style M1 fill:#FDEBD0,stroke:none,color:#000000,font-size:16px
    style M2 fill:#FDEBD0,stroke:none,color:#000000,font-size:16px
    style Y fill:#D5F5E3,stroke:none,color:#000000,font-size:16px
```

Recall that we use the DAGs to guide our study design and our analysis. A well-defined DAG tells us what questions are appropriate given the causal assumptions of our DAG. Generally, this causal DAG implies the following:

* the total effect of `X` on `Y` should adjust for confounders of `X -> Y`,
* mediators `M1` and `M2` should *not* be included in the total-effect (TE) model
* causal decomposition needs to be derived from the full SCM (again, this will be discussed in the next tutorial section)

Let us first address a simple way to model the total effect.

### Total effect

The total effect can be thought of as a primary driver in your system. This includes **all directed paths** from `X` to `Y`; it is the overall effect that you want to know about, such as: 

* Does temperature impact abundance?
* Does precipitation impact abundance?

In a kitchen-sink model, we may be tempted to throw both `temperature` and `precipitation` into the model and then compare coefficients. By now, we know that this approach could introduce bias. In the SCM world, the true total causal effect (TE) is assessed more directly, by comparing the expected outcome under two interventions (i.e. choices of `temperature` or `precipitation` values) on `X`. Let us see this in mathematical form:

$$
TE(x_1, x_0) = E[Y \mid do(X=x_1)] - E[Y \mid do(X=x_0)].
$$
This equation may have a new operator that you have never seen before. The `do` operator was created by Judea Pearl to indicate an explicit intervention. Armed with this knowledge, see if you can think about the elements of this equation. Think about what the equation is describing. Here are some reminders: 

* `*E*` = expected value (essentially, the average)
* `|` = read this as "given" (as in "value of `Y` given `X`)

Let us walk through the process of estimating a total effect from our set of causal assumptions.

## Workspace set-up

First, make sure that the following packages are loaded or added to your `load_packages.r` script.

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(mgcv)
library(dagitty)
library(igraph)
library(broom)
library(furrr)
library(future)
library(gratia)
library(DHARMa)
```

Let us know walk through the steps for estimating the total effect. 

## Step 0: Simulate a dataset to use for the SCM

We now simulate a dataset whose true generating process matches the DAG. The simulated example includes:

* one confounder `C`,
* one exposure `X`,
* two mediators `M1` and `M2`,
* a continuous outcome `Y`,
* two crossed random effects: `site` and `observer`,
* both linear and nonlinear edges.

The nonlinear edges will be:

* `X -> M1`
* `M2 -> Y`

```{r}
set.seed(20260408)

n_site <- 4
n_obs  <- 5
n_rep  <- 10
n      <- n_site * n_obs * n_rep

site <- factor(rep(paste0("site_", seq_len(n_site)), each = n_obs * n_rep))
obs  <- factor(rep(rep(paste0("obs_", seq_len(n_obs)), each = n_rep), times = n_site))

# crossed random intercepts
b_site_x  <- rnorm(n_site, 0, 0.55)
b_obs_x   <- rnorm(n_obs, 0, 0.45)

b_site_m1 <- rnorm(n_site, 0, 0.35)
b_obs_m1  <- rnorm(n_obs, 0, 0.30)

b_site_m2 <- rnorm(n_site, 0, 0.35)
b_obs_m2  <- rnorm(n_obs, 0, 0.30)

b_site_y  <- rnorm(n_site, 0, 0.70)
b_obs_y   <- rnorm(n_obs, 0, 0.55)

# exogenous confounder
C <- rnorm(n, mean = 0, sd = 1.1)

# exposure
X <- 0.85 * C +
  b_site_x[site] +
  b_obs_x[obs] +
  rnorm(n, 0, 1.0)

# mediator 1
M1 <- 1.0 * sin(X / 1.3) +
  0.55 * C +
  b_site_m1[site] +
  b_obs_m1[obs] +
  rnorm(n, 0, 0.8)

# mediator 2
M2 <- 0.60 * X +
  0.85 * M1 +
  0.18 * C +
  b_site_m2[site] +
  b_obs_m2[obs] +
  rnorm(n, 0, 0.9)

# outcome
Y <- 1.15 * X +
  1.10 * M2 -
  0.42 * (M2^2) +
  0.70 * C +
  b_site_y[site] +
  b_obs_y[obs] +
  rt(n, df = 6) * 0.9

dat <- tibble(
  site = site,
  obs  = obs,
  C    = C,
  X    = X,
  M1   = M1,
  M2   = M2,
  Y    = Y
)

glimpse(dat)
summary(dat$Y)
```

We purposefully added some non-linearities into this simulated datasets. Let us quickly inspect them to see what we have. 

```{r}
p_x_m1 <- ggplot(dat, aes(X, M1)) +
  geom_point(alpha = 0.08) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 10)) +
  labs(title = "Simulated nonlinear relation: X -> M1")

p_m2_y <- ggplot(dat, aes(M2, Y)) +
  geom_point(alpha = 0.08) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 10)) +
  labs(title = "Simulated nonlinear relation: M2 -> Y")

p_x_m1
p_m2_y
```


## Step 1: Encode the DAG and recover the appropriate adjustment set using `dagitty` in R

Recall that the adjustment set for the total effect should be derived from the DAG; it should not come from guesswork or intuition. We have explored how to use the online user interface for [`dagitty` ](https://dagitty.net/dags.html), but this can also be automated to some degree directly in R. Examine the following code and see how the DAG can be input into R using a specific syntax that can be manually prepared or copied from the right-hand code box in the `dagitty` user interface.

First, save your DAG as an object (named something like `dag_txt`):
```{r}
dag_txt <- dagitty("dag {
  C -> X
  C -> M1
  C -> Y
  X -> M1
  X -> M2
  X -> Y
  M1 -> M2
  M2 -> Y
}")
```

This formalizes the set of casual assumptions of your system. The next step is to specify the exposure and outcome variables that are specific to your question. In this case, your question asks about the total-effect of `X` on `Y`. So, set your exposure and outcome appropriately in the `adjustmentSets` function:

```{r}
dag_txt
adjustmentSets(dag_txt, exposure = "X", outcome = "Y")

```

For this DAG, `dagitty` should identify `C` as the minimal sufficient adjustment set for the total effect of `X` on `Y`. This is exactly what we expect for a minimal model for this generally canonical causal DAG. That is, for this the total-effect model, the model should:

* include `C` (the confounder) as a covariate (since this effectively blocks the other causal pathway to `Y`)
* not include mediators (`M1` or `M2`) as this would eliminate the indirect causal pathway thereby yielding the direct versus total-effect.

## Step 2: Fit the total-effect model using the DAG-derived adjustment set

Because the goal here is the total effect of `X` on `Y`, the DAG (and the `adjustmentSets` function) tells us to adjust for `C` and exclude mediators. That is simple enough. But we also need to include the two crossed random effects because repeated observations from the same site and observer are not independent. And, as we technically are naive about the dataset because we just collected it (and did not simulate it), we understand that nonlinear relationships might be possible. So, how do we determine the form of the total effects model? This will be the same logic and approach that you can use when conducting causal decomposition (though you might use slightly different tools). 

The answer lies in our old friend, the Akaike Information Criterion (AIC).

### Using AIC values for model comparison

AIC is useful for comparing candidate statistical representations of the same causal estimand. Specificaly, AIC can *only* be used to:

1. compare plausible error structures
2. compare plausible random-effects structures for the same fixed-effect specification (which is determined by your DAG logic)

AIC should not be used to: 

* decide the causal estimand (what is included in your model)
* decide the adjustment set (again, what to include in your model)

For AIC comparisons involving optimization random-effect structures, we use `method = "REML"`. After selecting random-effects structure based on lowest AIC, compare models with different forms (linear or nonlinear) using `method = "ML"`. 

Let us do that now!

### Candidate total-effect models

First, focus on the random effect structure

```{r}
# no random effect
mod_total_gaussian_none <- gam(
  Y ~ X + C,
  data = dat,
  family = gaussian(),
  method = "REML"
)

mod_total_gaussian_site <- gam(
  Y ~ X + C + s(site, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

mod_total_gaussian_obs <- gam(
  Y ~ X + C + s(obs, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

mod_total_gaussian_crossed <- gam(
  Y ~ X + C + s(site, bs = "re") + s(obs, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "REML"
)

```

Then, we compare using AIC. 

```{r}
aic_total_tbl <- tibble(
  model = c(
    "Gaussian, no random effects",
    "Gaussian, site random effect",
    "Gaussian, observer random effect",
    "Gaussian, crossed site + observer random effects"
  ),
  AIC = c(
    AIC(mod_total_gaussian_none),
    AIC(mod_total_gaussian_site),
    AIC(mod_total_gaussian_obs),
    AIC(mod_total_gaussian_crossed)
  )
) |>
  arrange(AIC) |>
  mutate(delta_AIC = AIC - min(AIC))

aic_total_tbl
```

Here, you can see that you have a choice. Does the more complex random effect structure reflect your study design more, or not? For now, let us assume that the "Gaussian, site random effect" model has the preferred random effects structure. We can now examine the functional form using `method = "ML"`.

```{r}
mod_total_gaussian_site <- gam(
  Y ~ X + C + s(site, bs = "re"),
  data = dat,
  family = gaussian(),
  method = "ML"
)
mod_total_scat_site <- gam(
  Y ~ X + C + s(site, bs = "re") ,
  data = dat,
  family = scat(),
  method = "ML"
)
```

Let us again compare using AIC, this time focusing only on the two models for which we wanted to compare the functional form (specifically, the error distribution)

```{r}
aic_total_tbl <- tibble(
  model = c(
    "Gaussian, site random effect",
    "Scaled-t, site random effect"

  ),
  AIC = c(
    AIC(mod_total_gaussian_site),
    AIC(mod_total_scat_site)
  )
) |>
  arrange(AIC) |>
  mutate(delta_AIC = AIC - min(AIC))

aic_total_tbl
```

What to do from here? Think about the following:

* the model with the smallest AIC is the preferred statistical representation among these candidates
* if two models are very close, prefer the simpler model unless the added complexity is justified

In the above case, the model with the scaled-t error distribution is the best of the candidate models. So, we can use that one for the final total-effects model.

### Get the final model

```{r}
mod_total <- gam(
  Y ~ X + C + s(site, bs = "re"),
  data = dat,
  family = scat(),
  method = "REML"
)

summary(mod_total)
plot(mod_total, all.terms=T, pages=1)

```

Now, for GAMs, checking the residual structure is a bit tricky. We must manually construct the residuals in `DHARMa` using a simulation. Here is how we do it.

```{r}
# simulate from model
sim <- simulate(mod_total, nsim = 1000)


# create DHARMa object manually
res <- createDHARMa(
  simulatedResponse = sim,
  observedResponse = dat$Y,
  fittedPredictedResponse = predict(mod_total, type = "response")
)

plot(res)

```

So, it not perfect and still has some structural issues. But we are close!

```{r}
m_linear <- gam(
  Y ~ X + C + s(site, bs = "re"),
  data = dat,
  family = scat(),
  method = "REML"
)

m_x <- gam(
  Y ~ s(X) + C + s(site, bs = "re"),
  data = dat,
  family = scat(),
  method = "REML"
)


m_c <- gam(
  Y ~ X + s(C) + s(site, bs = "re"),
  data = dat,
  family = scat(),
  method = "REML"
)

m_b <- gam(
  Y ~ s(X) + s(C) + s(site, bs = "re"),
  data = dat,
  family = scat(),
  method = "REML"
)


```

Check the AIC.


```{r}
aic_total_tbl <- tibble(
  model = c(
    "all linear",
    "smoothed X",
    "smoothed C",
    "smoothed both"

  ),
  AIC = c(
    AIC(m_linear),
    AIC(m_x),
    AIC(m_c),
    AIC(m_b)
  )
) |>
  arrange(AIC) |>
  mutate(delta_AIC = AIC - min(AIC))

aic_total_tbl
```

Simulate residuals

```{r}
# simulate from model
sim <- simulate(m_b, nsim = 1000)


# create DHARMa object manually
res <- createDHARMa(
  simulatedResponse = sim,
  observedResponse = dat$Y,
  fittedPredictedResponse = predict(m_b, type = "response")
)

plot(res)

```


## Next Up!

In the next section, we will conduct a formal causal decomposition!




