---
title: "GLM R Walkthrough: Puffer Jackalope Abundance"
format: html
execute:
  echo: true
  warning: false
  message: false
  error: false
editor: visual
editor_options:
  chunk_output_type: console
---

## Load packages

```{r}
source("chapters/class_r_scripts/load_packages.r")
```

## Import and inspect the data

```{r}
pufferi <- read.csv("chapters/class_exercise_data/lepusantilocapra_pufferi_abundance.csv")
str(pufferi) # examine data structure
```

## Plot abundance against depth by period

```{r}
p <- ggplot2::ggplot(pufferi, aes(x = mean_depth, y = abund)) +
  geom_point(color = "black", size = 3) +
  facet_wrap(~ period, ncol = 1, nrow = 2) +
  labs(x = "mean depth (km)", y = "number of _L. pufferi_") +
  theme_bw()
p
```

## Fit a baseline Gaussian GLM

```{r}
base_model <- glm(formula = abund ~ mean_depth, data = pufferi)
```

## Examine the model output

```{r}
summary(base_model) 
```

```{r}
# Tidy model coefficients
tidy_glm <- tidy(base_model, conf.int = TRUE)
print(tidy_glm)
```

```{r}
# Glance at model-level statistics
glance_glm <- glance(base_model)
print(glance_glm)
```

```{r}
# Augment the original data with predictions and residuals
augment_glm <- augment(base_model)
print(augment_glm)
```

## Diagnose the Gaussian GLM

```{r}
sim_res <- DHARMa::simulateResiduals(base_model)
```

```{r}
plot(sim_res)
```

## Check residuals against predictors and sampling effort

```{r echo=FALSE}
plotResiduals(sim_res, pufferi$mean_depth)
```

```{r}
plotResiduals(sim_res, pufferi$period)
```

```{r}
plotResiduals(sim_res, pufferi$year)
```

```{r}
plotResiduals(sim_res, pufferi$swept_area)
```

## Inspect the count response

```{r}
hist(pufferi$abund, breaks=100)
```

## Fit and diagnose a Poisson GLM

```{r}
base_model_pois <- glm(abund ~ mean_depth, 
  data = pufferi, 
  family = poisson(link = "log"))
```

```{r}
sim_res <- simulateResiduals(base_model_pois)
```

```{r}
plot(sim_res)
```

## Test for overdispersion

```{r}
testDispersion(sim_res)
```

## Fit a negative binomial GLM

```{r}
mod_nb <- glm.nb(abund ~ mean_depth, data = pufferi)
summary(mod_nb)
```

## Diagnose the negative binomial GLM

```{r}
testDispersion(mod_nb)
```

```{r}
sim_res <- simulateResiduals(mod_nb)
plot(sim_res)
```

## Check residuals against the other variables

```{r}
plotResiduals(sim_res, pufferi$mean_depth)
```

```{r}
plotResiduals(sim_res, pufferi$period)
```

```{r}
plotResiduals(sim_res, pufferi$year)
```

```{r}
plotResiduals(sim_res, pufferi$swept_area)
```

## Account for swept area with an offset

```{r}
pufferi$log_sa <- log(pufferi$swept_area)  # log-transform swept_area
mod_nb_offset <- glm.nb(abund ~ mean_depth + offset(log_sa), 
  data = pufferi)
```

## Diagnose the offset model

```{r}
sim_res <- simulateResiduals(mod_nb_offset)
```

```{r}
plotResiduals(sim_res, 
              pufferi$mean_depth)
```

```{r}
plotResiduals(sim_res, 
              pufferi$period)
```

```{r}
plotResiduals(sim_res, 
              pufferi$year)
```

## Revisit the DAG and refit the offset model

```{r}
pufferi$log_sa <- log(pufferi$swept_area)  # log-transform swept_area
mod_nb_offset <- glm.nb(abund ~ mean_depth + offset(log_sa), 
  data = pufferi)
```

## Fit alternative candidate models for comparison

```{r}
mod_nb_offset_01 <- glm.nb(abund ~ offset(log_sa), 
  data = pufferi)
mod_nb_offset_02 <- glm.nb(abund ~ mean_depth + offset(log_sa), 
  data = pufferi)
mod_nb_offset_03 <- glm.nb(abund ~ period + offset(log_sa), 
  data = pufferi)
mod_nb_offset_04 <- glm.nb(abund ~ year + offset(log_sa), 
  data = pufferi)
mod_nb_offset_05 <- glm.nb(abund ~
  mean_depth + period + offset(log_sa), 
  data = pufferi)
mod_nb_offset_06 <- glm.nb(abund ~
  mean_depth + year + offset(log_sa), 
  data = pufferi)
mod_nb_offset_07 <- glm.nb(abund ~
  period + year + offset(log_sa), 
  data = pufferi)
mod_nb_offset_08 <- glm.nb(abund ~
  mean_depth * period + year + offset(log_sa), 
  data = pufferi)
mod_nb_offset_09 <- glm.nb(abund ~
  mean_depth * year + period + offset(log_sa), 
  data = pufferi)
mod_nb_offset_10 <- glm.nb(abund ~
  mean_depth + year * period + offset(log_sa), 
  data = pufferi)
```

## Compare candidate models using AIC

```{r}
aic_df <- AIC(
  mod_nb_offset_01,
  mod_nb_offset_02,
  mod_nb_offset_03,
  mod_nb_offset_04,
  mod_nb_offset_05,
  mod_nb_offset_06,
  mod_nb_offset_07,
  mod_nb_offset_08,
  mod_nb_offset_09,
  mod_nb_offset_10
  )
aic_df <- aic_df |>
  arrange(AIC)
```

## Diagnose the selected example model

```{r}
sim_res <- simulateResiduals(mod_nb_offset_10)
plot(sim_res)
```

## Check residuals against depth, period, and year

```{r}
plotResiduals(sim_res, pufferi$mean_depth)
```

```{r}
plotResiduals(sim_res, pufferi$period)
```

```{r}
plotResiduals(sim_res, pufferi$year)
```
