---
title: "Tutorial: GAMs"
format: html
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction

In this tutorial, we introduce how to model different datasets using Generalized Additive Models (GAMs). Later on, we will address how to add a random effect into these models (turning these into Generalized Additive Mixed Models, or GAMMs). 

## Example #1: Building a basic GAM

For this example, we will revisit a familiar dataset: a population of *Lepusantilocapra pufferi* (puffer jackalope) starting in the mid-1970s. You will need the following file (simply click to save):

-   [`lepusantilocapra_pufferi_abundance.csv`](../chapters/class_exercise_data/lepusantilocapra_pufferi_abundance.csv)

As a reminder, the data include the following variables:

-   `site`: where the individuals were sampled
-   `abund`: number of individuals detected at each location
-   `dens`: density of individuals (per square meter) at each sampling location
-   `mean_depth`: mean depth, in meters of the sampling location
-   `year`: year of the study
-   `period`: 1=before and 2= after introduction of *Pelicanus pufferivorus* (puffer-jackalope pelican)
-   `x_km`: longitude (decimal degrees) of sampling location
-   `y_km`: latitude (decimal degrees) of sampling location
-   `swept_area`: size in square meters of the sampling location (effort)

Though we examined the relationship between total *Lepusantilocapra pufferi* abundance and mean depth for each time period (before/after introduction of puffer-jackalope pelicans), we will now improve our modeling process. Here, we will simplify our model and use density (`dens`) as our response variable (to account for variation in area searched, rather than using an offset). 

## Set up workspace
```{r message=FALSE, warning=FALSE}
source("class_r_scripts/load_packages.r")
```

Then read in the dataset.

```{r}
pufferi <- read.csv("class_exercise_data/lepusantilocapra_pufferi_abundance.csv")

```

## A linear model of jackalope density

```{r}
mod_gamma <- glmmTMB(dens ~
  mean_depth + year * period, family=Gamma(link="log"),
  data = pufferi)
```

What happened here do you think? Look at the `dens` variable. Seriously, stop and think about the original dataset and remember that `dens` is a metric derived by `abund` divided by `swept_area`. The values for `abund` is very small relative to the large `swept_area` square meters.

Let us try to solve this problem by converting the `swept_area` values into hectares (10000 square meters) and see if the linear model converges:

```{r}

# convert swept_area to hectares
pufferi$dens2 <- pufferi$abund / (pufferi$swept_area/10000)

mod_gamma2<- glmmTMB(dens2 ~
  mean_depth + year * period, family=Gamma(link="log"),
  data = pufferi)
```

So, the fact that the model onverged and did not throw other warning messages is promising. So, what happened? The original model was having issues (and throwing an NaN error) because one of the Gamma parameters could not easily be estimated due to the very small positive values in the original `dens` variable. 

Let us go ahead and simulate the residuals of this model using the `DHARMa` package (our usual approach).

```{r}
res <- simulateResiduals(mod_gamma2)
plot(res)
```
We have solved our main issue! But can this model be improved?

## Fitting a GAM 

First, let's convert our `glmmTMB` model to a model formula that can be used in the `mgcv::gam` function. Actually, there is nothing really to convert, as it looks identical.

```{r}
mod_gam<- gam(dens2 ~
  mean_depth + year * period, family=Gamma(link="log"),
  data = pufferi)
```
To convince you that it's doing the same thing, check out the summary of `mod_gam` and `mod_gamma2`.

```{r}
summary(mod_gamma2)

summary(mod_gam)
```
Very close indeed. But let us now see if we can improve this model (even in a small way). We are going to add a smoother to our one continuous variable, `mean_depth` and then refit the model.

```{r}
mod_gam<- gam(dens2 ~
  s(mean_depth, bs="cr", k=10) + year * period, family=Gamma(link="log"),
  data = pufferi)
```
This output looks a bit different. You now have two components to the model output: one for the parametric (linear) terms and one for the semiparametric (smoothed, or non-linear terms). We will discuss how to interpret the semiparametric effects in class. 

For the sake of a quick comparison (i.e. not what you really should do for a manuscript), we can examine the AICs of these models. 
```{r}
AIC(mod_gamma2, mod_gam)
```

Clearly, inclusion of the non-linear term has improved our model fit. Not dramatically so, but you get the point. 

### What about those knots (k) and the risk of overfitting?

As you recall from the introductory material, smoothed terms base their "smoothiness" or "curviness" on the number of segments (knots) that the line is split into. The more knots means that the relationship is more curvy. The various GAM functions and packages use slightly different methods for finding a smooth that does not overfit the data. But let's look at this ourselves in the next example!


## Example #2: Rainfall at a long term jackalope study site

To do this, let us look at some weather data! This is another dataset you will need to access (simply click to save):

-   [`rainfall_daily.csv`](../chapters/class_exercise_data/rainfall_daily.csv)

This dataset contains daily rainfall data across 83 years on Barro Colorado Island, the home of the endemic liana jackalope (_Lepusantilocapra cryptophiloliana_). We want to understand recent temporal patterns (among and within years) in daily rainfall. So, let us read in the data and then, because we are interested in recent years only--reduce its size a bit (by limiting to the year range of 2008-2011). We will also go ahead and ensure the dates are in the right format and use these to calculate a a derived variable for data difference (relative to the first date in new, filtered dataset).

```{r}
rain <- read_csv("class_exercise_data/rainfall_daily.csv")

rain <- rain |> filter(year > 1990)

rain <- rain |> 
  mutate(
    date2 = ymd(date),  # Convert date to Date format using lubridate's ymd
    yearday2 = yearday + 1,  # Increment yearday by 1
    days_diff = as.numeric(date2 - min(date2, na.rm = TRUE)) # Diff from min date
  )
```
Let us plot these data!
```{r}
# Plot rainfall with shading for certain year ranges
rain_sorted <- rain |> arrange(yearday2)

# Line plot for pooled data across all years
ggplot(rain_sorted, aes(x = yearday, y = rainfall_mm)) +
  geom_line(color = "darkblue", linewidth = 0.8) +  
  labs(title = "Pooled, All Years", x = "Day of Year", y = "Rainfall (mm)") +
  theme_minimal()
```

The first thing we notice is that there is a distinct dry season in the first part of the calendar year. Then, after about 100 days or so (corresponding to sometime in April), the rainy season begins. 

### Modeling rainfall as a function of day of year

We then move into modeling rainfall. First, we need to understand the boundaries of our response variable:

```{r}
range(rain$rainfall_mm)
hist(rain$rainfall_mm, 50)
```

The distribution includes 0 values and has a long right tail. A Gamma will work very well here, but Gamma distributions cannot include 0 values. So, we can do a tricky little adjustment by adding a small value to all 0 values:

```{r}
rain <- rain %>%
  mutate(rainfall_mm = if_else(rainfall_mm == 0,
                               0.00001,
                               rainfall_mm))

```

Now, we can fit GAMs with `rainfall_mm` as the response variable and a smoother of `yearday`. We will specify a Gamma error distribution with a _log_ link function. But let us not stop there. Let us run several versions of these GAMs each with a different number of knots (3-20). 

```{r}
mod3 <- gam(rainfall_mm ~ s(yearday, k=3, bs="cc"), family=Gamma(link="log"), data=rain)
mod4 <- gam(rainfall_mm ~ s(yearday, k=4, bs="cc"), family=Gamma(link="log"), data=rain)
mod5 <- gam(rainfall_mm ~ s(yearday, k=5, bs="cc"), family=Gamma(link="log"), data=rain)
mod6 <- gam(rainfall_mm ~ s(yearday, k=6, bs="cc"), family=Gamma(link="log"), data=rain)
mod8 <- gam(rainfall_mm ~ s(yearday, k=8, bs="cc"), family=Gamma(link="log"), data=rain)
mod10 <- gam(rainfall_mm ~ s(yearday, k=10, bs="cc"), family=Gamma(link="log"), data=rain)
mod15 <- gam(rainfall_mm ~ s(yearday, k=15, bs="cc"), family=Gamma(link="log"), data=rain)
mod20 <- gam(rainfall_mm ~ s(yearday, k=20, bs="cc"), family=Gamma(link="log"), data=rain)
```

Then, we can compare using AIC.

```{r}
AIC(mod3, mod4, mod5, mod6, mod8, mod10, mod15, mod20)
```

What do we see?

```{r}
par(mfrow=c(3, 3), mar=c(1,1,1,1))
plot(mod3, shade=T)
plot(mod4, shade=T)
plot(mod5, shade=T)
plot(mod6, shade=T)
plot(mod8, shade=T)
plot(mod10, shade=T)
plot(mod15, shade=T)
plot(mod20, shade=T)
```

Some would say that this is a classic example of the dangers of overfitting, which GAMs may have a tendency to do or smaller sample sizes. Sometimes, the overfitting may be useful (for finding potential environmental thresholds). But, generally, this example highlights the balance between _wiggliness_ and fit. You will find that many practitioners conservatively choose five knots just to be safe; there is currently no statistical justification for this choice.

### Modeling rainfall as a function of day of year

Now, let us add complexity. We shall add a term of `year` so that we can separate the long-term trends in rainfall (across years) from the within-year variation. 

We therefore fit the model with two main effect, `yearday` and `year`. For yearday, we want out final predicted output to recognize the seamlessness in our data, namely that the day of the year _after_ the 365th day is the 1st day. Therefore, we specify a cyclic cubic regression spline using the appropriate basis function`bs = "cc"`. We set a large maximum number of knots (n=10) for illustration purposes only.

```{r}
rain_mod_01 <- gam(
  rainfall_mm ~ 
    s(yearday, bs = "cc", k = 10) +      # average seasonal curve
    s(year, k=10),                     # long-term trend
  family = Gamma(link = "log"),
  data = rain,
  method = "REML"
)

```
We can then either look at the `summary` or plot the conditional effects of the model terms. 

```{r}
rain_mod <- gam(
  rainfall_mm ~ 
    s(yearday, bs = "cc", k = 10) +      # average seasonal curve
    s(year, k=10) +                     # long-term trend
    ti(yearday, year, 
       bs = c("cc", "tp"), 
       k = c(10, 10)),                    # year-specific seasonal deviations
  family = Gamma(link = "log"),
  data = rain,
  method = "REML"
)

summary(rain_mod)

```

